### 2021-06-10 paper/report.tex

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination-induced herd immunity
\citep{Swaminathan2021}.\comment[id=K]{WHO Press Conference saying we won't reach herd
immunity in 2021. She cites no source.}\comment[id=HM]{If we find a paper (s)he cites, would be even better}
In the meantime, it is of utmost importance to employ an effective mix of strategies for
containing the virus. The most frequent initial response was a set of non-pharmaceutical
interventions (NPIs) to reduce contacts between individuals. While this has allowed some
countries to sustain equilibria with very low infection numbers\footnote{See
    \citet{Contreras2021} for a theoretical equilibrium at low case numbers which is
    sustained with test-trace-and-isolate policies.}, most have seen large fluctuations of
infection rates. Containment measures have become increasingly diverse and now include
testing, more nuanced NPIs, and contact tracing. Neither these policies' effect nor the
influence of seasonal patterns or more infectious virus strains are well understood in
quantitative terms. This paper develops a model incorporating all these factors. The
framework allows to combine a wide variety of data in a timely fashion, making it useful
to predict the effects of various interventions. We apply the model to Germany and show
that rapid testing had the largest impact on the reduction in infections by almost 80\%
during the month of May 2021. We conclude that rapid tests have a large role to play at
least as long as vaccinations have not been offered to an entire population.

At the core of our agent-based model are physical contacts between heterogeneous agents
(Figure~\ref{fig:broad_model_description}).\footnote{We provide a detailed comparison to
    other approaches in \ref{sec:literature_review}. The model most closely related to ours
    is described in \citet{Hinch2020}.} Each contact between an infectious individual and
somebody susceptible to the disease bears the risk of transmitting the virus. Contacts
occur in the household, at work, at school, or in other settings (leisure activities,
grocery shopping, medical appointments, etc.). Some contacts recur regularly, others
occur at random. Random contacts are typically assortative in age and geographical
location. Empirical applications can take the population structure from census data and
the types and frequency of contacts from diary data measuring contacts before the
pandemic \citep[e.g.][]{Mossong2008}.\footnote{\citet{Hoang2019} provide access to
    multiple data sets on contact types and frequencies at
    \url{http://www.socialcontactdata.org/} covering countries from all continents except
    North America and Australia.} The dimensions are chosen so that the most common NPIs can
be modeled in great detail by reducing the number of contacts in a particular setting or
the risk of transmitting the disease for a type of contact. For example, a mandate to
work from home will reduce the number of work contacts to zero for a fraction of the
working population. Schools and daycare can be closed entirely, operate at reduced
capacity---including an alternating schedule---,or implement mitigation measures like
masking requirements or air filters \citep{Lessler2021}. Curfews may reduce the number
of contacts in non-work/non-school settings. In any setting, measures like masking
requirements would reduce the probability of infection associated with a contact.

@@

We include two types of tests. Polymerase chain reaction (PCR) tests directly reveal
whether an individual is infected or not. PCR tests require some days to be processed
and there are always aggregate capacity constraints. In contrast, rapid antigen tests
yield immediate results. Specificity and sensitivity of these tests is set according to
data analyzed in \cite{Bruemmer2021, Smith2021}; sensitivity depends on the timing of
the test relative to the start of infectiousness. Figure~\ref{fig:antigen_tests} shows
our model for rapid test demand. Schools may require students to be tested regularly.
Rapid tests may be offered by employers for on-site workers. Individuals may demand
tests for private reasons, which include having plans to meet other people\footnote{A
    positive test will make them reduce their contacts; this is why tests impact the actual
    contacts in Figure~\ref{fig:model-description}.}, showing symptoms of CoViD-19, and
because a household member tested positively for the virus. We endow each agent with an
individual compliance parameter. This parameter determines whether she takes up rapid
tests offered by employers or follows up on private reasons. The thresholds are lower
for tests in a private setting than for tests at the workplace.\comment[id=HM]{True?
@Janos}

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and estimate the model's many
parameters.\footnote{See section S.XXX of the supplementary materials for an overview.}
Mobility data is used to model the reductions in work contacts. School and daycare
policies are incorporated directly from official directives. Infection rates by age and
geographical region are estimated to match to officially recorded numbers; so is the
prevalence of virus strains. The model yields total infections, but only a fraction of
those will be officially recorded. We model the fraction of known cases as depicted in
Figure~\ref{fig:model_for_official_cases}.\comment[id=HM]{Add a sentence.}
% \footnote{EQUATIONS FOR THE FOURTH PICTURE (find names!):

%     The probabilities for Figure~\subref{fig:model_for_official_cases} are the
%     following:

%     \begin{align*} x & = \frac{n\_newly\_infected \times share\_known\_cases \times
%         P(PCR | S)}{n\_symptomatic\_and\_untested}                    \\
%         y & = \frac{n\_newly\_infected \times share\_known\_cases \times P(PCR |
%         S)}{n\_asymptoptomatic\_and\_infected\_and\_untested} \\
%         z & = \frac{P(PCR | RT)}{survey} \end{align*} where untested means that
%     individuals are not waiting on the result of a previously administered rapid
%     test.} 
A further advantage is that the simulated data have a structure that resembles datasets
used for regression models, which allows additional plausibility checks by re-running
the same models on the model-generated data.\comment[id=HM]{Only keep if we actually do
something like that.}

We apply this model to Germany. In March and April 2020, the country broke the first
wave of the pandemic fairly quickly. Between May and mid-September, daily new infections
were below 20 per Million and day \citep{owidcoronavirus}. We model the period
mid-September 2020 to the end of May 2021. We pick the starting date for two reasons.
First, we do not include the first wave because the environment was very different
(e.g., aggregate PCR test capacity was much lower and we would require a very different
model for calculating the share of known cases). Second, a large fraction of cases
during summer of 2020 were traced to international travel
\citep{KochInstitut2021,Hodcroft2021} but the precise number is difficult to model.

Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the black line in Figure~\ref{fig:stringency_index} the Oxford Response
Stringency Index \citep{Hale2020}, which tracks the tightness of non-pharmaceutical
interventions. We transform the index so that lower values represent higher levels of
restrictions. A value of zero means all measures incorporated in the index are turned
on. The value 1 represents the situation in mid-September, with restrictions on
gatherings and public events, masking requirements, but open schools and workplaces (the
raw value of the index at that point is 49.5). In the seven weeks between mid September
and early November, cases increased by a factor of 10. Restrictions were somewhat
tightened in mid-October and again in early November. New infections remained constant
throughout November, before rising again in December, which prompted the most stringent
lockdown to this date. Schools and daycare centers were closed again, so were
customer-facing businesses except for grocery and drug stores. From the peak of the
second wave just before Christmas until the trough in mid-February, newly detected cases
decreased by almost three quarters. The third wave in the spring of 2021 is associated
with the B.1.1.7 strain, which became dominant in March. See
Figure~\ref{fig:share_b117}. In early March, some NPIs were being relaxed; e.g.,
hairdressers and home improvement stores were allowed to open again to the public.

@@

By this time, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical personnel (Figure~\ref{fig:antigen_tests_vaccinations}. By the end of May, just
over 40\% had received at least one dose of a vaccine. Around the same time, rapid tests
started to replace PCR tests in many medical and nursing facilities. These had to be
administered by medical doctors or in pharmacies. At-home tests approved by authorities
became available in mid-March, rapid test centers were opened and one test per person
and week was made available free of charge. Depending on the state, customers were only
allowed to enter certain stores with a recent negative rapid test result. These
developments are characteristic of many countries: The initial focus on NPIs to slow the
spread of the disease has been accompanied by vaccines and a growing acceptance and use
of rapid tests. At broadly similar points in time, novel strains of the virus have
started to pose additional challenges.

Our model is able to track these developments very well. The blue line in
Figure~\ref{fig:aggregated_fit} shows our model's predictions are very close to
officially recorded cases in the aggregate. This is also true for infections by age and
geographical region, which are shown in the supplementary materials
(Figures~\ref{fig:age_group_fit} and \ref{fig:state_fit}, respectively). We can
disentangle various mechanisms due to the distinct temporal variation in the drivers of
the pandemic. Next to the stringency index, the three lines in
Figure~\ref{fig:stringency_index} summarize how contact reductions, increased hygiene
regulations, and seasonality evolved since early September for each of the three broad
contact networks. For example, a value of 0.75 for the work multiplier means that if the
environment was the same as in September (levels of infection rates, no rapid tests or
vaccinations, only the wildtype virus present), infections at the workplace would be
reduced by 25\%. This reduction is the product of the effect of contact reductions,
increased hygiene regulations, and seasonality. Along with the levels of infections,
these measures determine the spread of SARS-CoV-2 in 2020. See
Appendix~\comment[id=HM]{Reference} for separate plots of the three factors. Two aspects
are particularly interesting. First, all lines broadly follow the stringency index and
they would do so even more if we left out seasonality and school vacations (roughly the
last two weeks of October, two weeks each around Christmas and Easter, and some days in
late May). Second, the most stringent regulations are associated with the period of
strong decreases in new infection between late December 2020 and mid-February 2021. The
measures were not enough, however, to stop the B.1.1.7 variant from spreading in the
subsequent period. The steep drop in recorded cases during May 2021 is associated with
at least weekly rapid tests to around 42~percent of the population, a vaccination rate
that rose from 28\% to 43\%, and mostly seasonality impacting a fall in the relative
infectiousness of contacts outside of work and school.

Figure~\ref{fig:2021_scenarios_broad} consider the relative effects of rapid tests,
vaccinations, and of seasonality during 2021, assuming NPIs to have evolved the same way
as in the baseline scenario. Figure~\ref{fig:2021_scenarios_recorded} shows the model
fit (the blue line, same as in Figure~\ref{fig:aggregated_fit}), a scenario without any
of the three factors (red line), and three scenarios turning these factors off one by
one. Figure~\ref{fig:2021_scenarios_newly_infected} does the same for total infections
in the model. Figure~\ref{fig:2021_scenarios_decomposition} employs Shapley values to
decompose the difference in total infections between the scenario without any of the
three factors and our main specification.

>>>>

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination-induced herd immunity
\citep{Swaminathan2021}.\comment[id=K]{WHO Press Conference saying we won't reach herd
immunity in 2021. She cites no source.}\comment[id=HM]{If we find a paper (s)he cites, would be even better}
In the meantime, it is of utmost importance to employ an effective mix of strategies for
containing the virus. The most frequent initial response was a set of non-pharmaceutical
interventions (NPIs) to reduce contacts between individuals. While this has allowed some
countries to sustain equilibria with very low infection numbers\footnote{See
    \citet{Contreras2021} for a theoretical equilibrium at low case numbers which is
    sustained with test-trace-and-isolate policies.}, most have seen large fluctuations of
infection rates. Containment measures have become increasingly diverse and now include
testing, more nuanced NPIs, and contact tracing. Neither these policies' effect nor the
influence of seasonal patterns or more infectious virus strains are well understood in
quantitative terms.

This paper develops a quantitative model incorporating all these factors. The framework
allows to combine a wide variety of data and mechanisms in a timely fashion, making it
useful to predict the effects of various interventions. We apply the model to Germany
and show that rapid testing had the largest impact on the reduction in infections by
almost 80\% during the month of May 2021. We conclude that rapid tests have a large role
to play at least as long as vaccinations have not been offered to an entire population.

At the core of our agent-based model are physical contacts between heterogeneous agents
(Figure~\ref{fig:broad_model_description}).\footnote{We provide a detailed comparison to
    other approaches in \ref{sec:literature_review}. The model most closely related to ours
    is described in \citet{Hinch2020}.} Each contact between an infectious individual and
somebody susceptible to the disease bears the risk of transmitting the virus. Contacts
occur in the household, at work, at school, or in other settings (leisure activities,
grocery shopping, medical appointments, etc.). Some contacts recur regularly, others
occur at random. Random contacts are typically assortative in age and geographical
location. Empirical applications can take the population structure from census data and
the types and frequency of contacts from diary data measuring contacts before the
pandemic \citep[e.g.][]{Mossong2008}.\footnote{\citet{Hoang2019} provide access to
multiple data sets on contact types and frequencies covering countries from four
continents.}

The three dimensions of contact types are chosen so that the most common NPIs can be
modeled in great detail. NPIs affect the number of contacts or the risk of transmitting
the disease upon having physical contact. The effect of different NPIs will generally
vary across contact types. For example, a mandate to work from home will reduce the
number of work contacts to zero for a fraction of the working population. Schools and
daycare can be closed entirely, operate at reduced capacity---including an alternating
schedule---,or implement mitigation measures like masking requirements or air filters
\citep{Lessler2021}. Curfews may reduce the number of contacts in non-work/non-school
settings. In any setting, measures like masking requirements would reduce the
probability of infection associated with a contact.

@@

We include two types of tests. Polymerase chain reaction (PCR) tests directly reveal
whether an individual is infected or not. PCR tests require some days to be processed
and there are aggregate capacity constraints throughout. In contrast, rapid antigen
tests yield immediate results; after a phase-in period, all tests that are demanded will
be performed. Specificity and sensitivity of these tests is set according to data
analyzed in \cite{Bruemmer2021, Smith2021}; sensitivity depends on the timing of the
test relative to the start of infectiousness. Figure~\ref{fig:antigen_tests} shows our
model for rapid test demand. Schools may require students to be tested regularly. Rapid
tests may be offered by employers for on-site workers. Individuals may demand tests for
private reasons, which include having plans to meet other people\footnote{A positive
test will make them reduce their contacts; this is why tests impact the actual contacts
in Figure~\ref{fig:model-description}.}, showing symptoms of CoViD-19, and because a
household member tested positively for the virus. We endow each agent with an individual
compliance parameter. This parameter determines whether she takes up rapid tests offered
by employers or follows up on private reasons. The thresholds are lower for tests in a
private setting than for tests at the workplace.\comment[id=HM]{True? @Janos}

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and estimate the model's many
parameters.\comment[id=HM]{
% \footnote{
See section S.XXX of the supplementary materials for an overview.
% }
} Mobility data is used to model the reductions in work contacts. School and daycare
policies can be incorporated directly from official directives. Aggregate tests,
vaccinations, and the prevalence of virus strains can be taken from official records and
from survey data.\comment[id=HM]{Anything else we need to mention?} We estimate the
infection probabilities\comment[id=HM]{need to be more precise here, cannot find in
Appendix} that the simulated new infections predict age- and region-specific time
series. However, the two sets of time series are not comparable directly because not all
cases will be recorded. We model the fraction of recorded cases as depicted in
Figure~\ref{fig:model_for_official_cases}.\comment[id=HM]{Add a sentence.}

The model is applied to the second and third wave of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021.
Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the black line in Figure~\ref{fig:stringency_index} the Oxford Response
Stringency Index \citep{Hale2020}, which tracks the tightness of non-pharmaceutical
interventions. We transform the index so that lower values represent higher levels of
restrictions. A value of zero means all measures incorporated in the index are turned
on. The value 1 represents the situation in mid-September, with restrictions on
gatherings and public events, masking requirements, but open schools and workplaces. In
the seven weeks between mid September and early November, cases increased by a factor of
10. Restrictions were somewhat tightened in mid-October and again in early November. New
infections remained constant throughout November, before rising again in December, which
prompted the most stringent lockdown to this date. Schools and daycare centers were
closed again, so were customer-facing businesses except for grocery and drug stores.
From the peak of the second wave just before Christmas until the trough in mid-February,
newly detected cases decreased by almost three quarters. The third wave in the spring of
2021 is associated with the B.1.1.7 strain, which became dominant in March. See
Figure~\ref{fig:share_b117}. In early March, some NPIs were being relaxed; e.g.,
hairdressers and home improvement stores were allowed to open again to the public. There
were many changes in details of regulations afterwards, but they did not change the
stringency index.

@@

By this time, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical personnel (Figure~\ref{fig:antigen_tests_vaccinations}. By the end of May, just
over 40\% had received at least one dose of a vaccine. Around the same time, rapid tests
started to replace regular PCR tests for staff in many medical and nursing facilities.
These had to be administered by medical doctors or in pharmacies. At-home tests approved
by authorities became available in mid-March, rapid test centers were opened and one
test per person and week was made available free of charge. Depending on the state,
customers were only allowed to enter certain stores with a recent negative rapid test
result. These developments are characteristic of many countries: The initial focus on
NPIs to slow the spread of the disease has been accompanied by vaccines and a growing
acceptance and use of rapid tests. At broadly similar points in time, novel strains of
the virus have started to pose additional challenges.

We draw simulated samples of agents from the distribution of recorded infections in
September 2020 and use the model to predict recorded infection rates until the end of
May 2021. See Supplementary Materials~\ref{sec:data_and_parameters} and \ref{sec:model}
for detailed descriptions of the data and the model, respectively. The blue line in
Figure~\ref{fig:aggregated_fit} shows our model's predictions are very close to
officially recorded cases in the aggregate. This is also true for infections by age and
geographical region, which are shown in the supplementary materials
(Figures~\ref{fig:age_group_fit} and \ref{fig:state_fit}, respectively).

The effects of various mechanisms can be disentangled due to the distinct temporal
variation in the drivers of the pandemic. Next to the stringency index, the three lines
in Figure~\ref{fig:stringency_index} summarize how contact reductions, increased hygiene
regulations, and seasonality evolved since early September for each of the three broad
contact networks. For example, a value of 0.75 for the work multiplier means that if the
environment was the same as in September (levels of infection rates, no rapid tests or
vaccinations, only the wildtype virus present), infections at the workplace would be
reduced by 25\%. The lines show the product of the effect of contact reductions,
increased hygiene regulations, and seasonality. See Appendix~\comment[id=HM]{Reference}
for separate plots of the three factors. Two aspects are particularly interesting.
First, all lines broadly follow the stringency index and they would do so even more if
we left out seasonality and school vacations (roughly the last two weeks of October, two
weeks each around Christmas and Easter, and some days in late May). Second, the most
stringent regulations are associated with the period of strong decreases in new
infections between late December 2020 and mid-February 2021. The reversal of the trend
is associated with he spread of the B.1.1.7 variant. The steep drop in recorded cases
during May 2021 is associated with at least weekly rapid tests to around 42~percent of
the population, a vaccination rate that rose from 28\% to 43\%, and mostly seasonality
impacting a fall in the relative infectiousness of contacts outside of work and school.

In order to better understand the contributios  of rapid tests, vaccinations, and of
seasonality on the evolution of infections in 2021,
Figure~\ref{fig:2021_scenarios_broad} considers various scenarios. NPIs are always the
same as in the baseline scenario. Figure~\ref{fig:2021_scenarios_recorded} shows the
model fit (the blue line, same as in Figure~\ref{fig:aggregated_fit}), a scenario
without any of the three factors (red line), and three scenarios turning these factors
off one by one. Figure~\ref{fig:2021_scenarios_newly_infected} does the same for total
infections in the model. Figure~\ref{fig:2021_scenarios_decomposition} employs Shapley
values to decompose the difference in total infections between the scenario without any
of the three factors and our main specification.
